\textbf{SPECFEM3D}\\ \textbf{Cartesian}\\ \textbf{User Manual}

### 2022-11-30 doc/USER_MANUAL/A_reference_frame.tex
updates tex files (\\ to \newline) for better pandoc conversions to wiki pages

The SPECFEM3D Cartesian software code internally uses Cartesian coordinates.
The given locations for sources and receiver locations thus may get
converted. Sources and receiver locations are read in from the \texttt{CMTSOLUTION}
(or \texttt{FORCESOLUTION}) and \texttt{STATIONS} files. Note that
e.g. the \texttt{CMTSOLUTION} and \texttt{FORCESOLUTION} files denotes
the location by \textquotedbl{}longitude/latitude/depth\textquotedbl{}.
We read in longitude as $x$ coordinate, latitude as $y$ coordinate.

In case the flag \texttt{SUPPRESS\_UTM\_PROJECTION} is set to \texttt{.false.}
in the main parameter file (see Chapter~\ref{cha:Creating-Distributed-Databases}),
the $x$/$y$ coordinates have to be given in degrees and are converted
to Cartesian coordinates using a UTM conversion with the specified
UTM zone.

The value for depth (given in $km$ in \texttt{CMTSOLUTION} and \texttt{FORCESOLUTION})
or burial depth (given in $m$ in \texttt{STATIONS}) is evaluated
with respect to the surface of the mesh at the specified $x$/$y$
location to find the corresponding $z$ coordinate. It is possible
to use this depth value directly as $z$ coordinate by changing the
flag \texttt{USE\_SOURCES\_RECEIVERS\_Z} to \texttt{.true.} in
\texttt{DATA/Par\_file}.

@@

The seismogram output directions are given in Cartesian $x$/$y$/$z$
directions and not rotated any further. Changing flags in \texttt{setup/constants.h} only rotates the seismogram
outputs if receivers are forced to be located at the surface (\texttt{RECEIVERS\_CAN\_BE\_BURIED}
set to \texttt{.false.}) and the normal to the surface at the receiver
location should be used (\texttt{EXTERNAL\_MESH\_RECEIVERS\_NORMAL} set to \texttt{.true.})
as vertical. In this case, the outputs are rotated to have the vertical
component normal to the surface of the mesh, $x$ and $y$ directions
are somewhat arbitrary orthogonal directions along the surface.

For the labeling of the seismogram channels, see Appendix~\ref{cha:channel-codes}.
Additionally, we add labels to the synthetics using the following
convention: For a receiver station located in an
\begin{description}
\item [{elastic domain:}] ~
\begin{itemize}
\item \texttt{semd} for the displacement vector
\item \texttt{semv} for the velocity vector
\item \texttt{sema} for the acceleration vector
\end{itemize}
\item [{acoustic domain:}] ~\\
 (please note that receiver stations in acoustic domains must be buried.
This is due to the free surface condition which enforces a zero pressure
at the surface.)
\begin{itemize}
\item \texttt{semd} for the displacement vector
\item \texttt{semv} for the velocity vector
\item \texttt{sema} for pressure which will be stored in the vertical component
\texttt{Z}. Note that pressure in the acoustic media is isotropic
and thus the pressure record would be the same in the other two component
directions. We therefore use the other two seismogram components to
store the acoustic potential in component \texttt{X} (or \texttt{N})
and the first time derivative of the acoustic potential in component
\texttt{Y} (or \texttt{E}).
\end{itemize}
\end{description}
The seismograms are by default written out in ASCII-format to the
\texttt{OUTPUT\_FILES/} subdirectory by each parallel process. You
can change this behavior by changing the following flags in the \texttt{DATA/Par\_file}:
\begin{description}
\item [{\texttt{USE\_BINARY\_FOR\_SEISMOGRAMS}}] set to \texttt{.true.} to have seismograms
written out in binary format.
\item [{\texttt{WRITE\_SEISMOGRAMS\_BY\_MAIN}}] Set to \texttt{.true.}
to have only the main process writing out seismograms. This can
be useful on a cluster, where only the main process node has access
to the output directory.
\item [{\texttt{USE\_OUTPUT\_FILES\_PATH}}] Set to \texttt{.false.} to
have the seismograms output to \texttt{LOCAL\_PATH} directory specified
in the main parameter file \texttt{DATA/Par\_file}.
\end{description}

>>>>

The SPECFEM3D Cartesian software code internally uses Cartesian coordinates.
The given locations for sources and receiver locations thus may get
converted. Sources and receiver locations are read in from the \texttt{CMTSOLUTION}
(or \texttt{FORCESOLUTION}) and \texttt{STATIONS} files. Note that
e.g. the \texttt{CMTSOLUTION} and \texttt{FORCESOLUTION} files denotes
the location by \textquotedbl{}longitude/latitude/depth\textquotedbl{}.
We read in longitude as $x$ coordinate, latitude as $y$ coordinate.\newline

In case the flag \texttt{SUPPRESS\_UTM\_PROJECTION} is set to \texttt{.false.}
in the main parameter file (see Chapter~\ref{cha:Creating-Distributed-Databases}),
the $x$/$y$ coordinates have to be given in degrees and are converted
to Cartesian coordinates using a UTM conversion with the specified
UTM zone.\newline

The value for depth (given in $km$ in \texttt{CMTSOLUTION} and \texttt{FORCESOLUTION})
or burial depth (given in $m$ in \texttt{STATIONS}) is evaluated
with respect to the surface of the mesh at the specified $x$/$y$
location to find the corresponding $z$ coordinate. It is possible
to use this depth value directly as $z$ coordinate by changing the
flag \texttt{USE\_SOURCES\_RECEIVERS\_Z} to \texttt{.true.} in
\texttt{DATA/Par\_file}.\newline

@@

The seismogram output directions are given in Cartesian $x$/$y$/$z$
directions and not rotated any further. Changing flags in \texttt{setup/constants.h} only rotates the seismogram
outputs if receivers are forced to be located at the surface (\texttt{RECEIVERS\_CAN\_BE\_BURIED}
set to \texttt{.false.}) and the normal to the surface at the receiver
location should be used (\texttt{EXTERNAL\_MESH\_RECEIVERS\_NORMAL} set to \texttt{.true.})
as vertical. In this case, the outputs are rotated to have the vertical
component normal to the surface of the mesh, $x$ and $y$ directions
are somewhat arbitrary orthogonal directions along the surface.\newline

For the labeling of the seismogram channels, see Appendix~\ref{cha:channel-codes}.
Additionally, we add labels to the synthetics using the following
convention: For a receiver station located in an
\begin{description}
\item [{elastic domain:}] ~
\begin{itemize}
\item \texttt{semd} for the displacement vector
\item \texttt{semv} for the velocity vector
\item \texttt{sema} for the acceleration vector
\end{itemize}
\item [{acoustic domain:}] ~\newline
 (please note that receiver stations in acoustic domains must be buried.
This is due to the free surface condition which enforces a zero pressure
at the surface.)
\begin{itemize}
\item \texttt{semd} for the displacement vector
\item \texttt{semv} for the velocity vector
\item \texttt{sema} for pressure which will be stored in the vertical component
\texttt{Z}. Note that pressure in the acoustic media is isotropic
and thus the pressure record would be the same in the other two component
directions. We therefore use the other two seismogram components to
store the acoustic potential in component \texttt{X} (or \texttt{N})
and the first time derivative of the acoustic potential in component
\texttt{Y} (or \texttt{E}).
\end{itemize}
\end{description}
The seismograms are by default written out in ASCII-format to the
\texttt{OUTPUT\_FILES/} subdirectory by each parallel process. You
can change this behavior by changing the following flags in the \texttt{DATA/Par\_file}:
\begin{description}
\item [{\texttt{USE\_BINARY\_FOR\_SEISMOGRAMS}}] set to \texttt{.true.} to have seismograms
written out in binary format.
\item [{\texttt{WRITE\_SEISMOGRAMS\_BY\_MAIN}}] Set to \texttt{.true.}
to have only the main process writing out seismograms. This can
be useful on a cluster, where only the main process node has access
to the output directory.
\item [{\texttt{USE\_OUTPUT\_FILES\_PATH}}] Set to \texttt{.false.} to
have the seismograms output to \texttt{LOCAL\_PATH} directory specified
in the main parameter file \texttt{DATA/Par\_file}.
\end{description}

### 2022-11-30 doc/USER_MANUAL/C_troubleshooting.tex

Examine the log file 'config.log'. It contains detailed informations.
In many cases, the paths to these specific compiler commands F90,
CC and MPIF90 will not be correct if `./configure` fails.

@@

Make sure you're pointing to the right 'mpif90' wrapper command.

Normally, this message will appear when you are mixing two different
Fortran compilers. That is, using e.g. gfortran to compile non-MPI
files and mpif90, wrapper provided for e.g. ifort, to compile MPI-files.

@@

which gives you that: 27,648 / 33,792 $\sim$ 81.8 \% of all elements
are not skewed, i.e. regular elements. a fantastic value :)

@@

about 6 \% of all elements have distortions with a skewness value
between 0.6 and 0.65. The skewness values give you a hint of how good
your mesh is. In an ideal world, you would want to have no distortions,
just like the 81\% from above. Those elements give you the best approximate
values by the GLL quadrature used in the spectral-element method.
However, having weakly distorted elements is still fine and the solutions
are still accurate enough. So empirically, values up to around 0.7
are tolerable, above that you should consider remeshing...

@@

You will see that the mesh contains a doubling layer, where we stitch
elements together such that the size of two elements will transition
to the size of one element (very useful to keep the ratio of wavespeed
/ element\_size about constant). Those elements in this doubling layer
have higher skewness values and make up those 6 \% in the histogram.

\item [{the code gives following error message "need at least one receiver":}] This
means that no stations given in the input file \texttt{DATA/STATIONS}
could be located within the dimensions of the mesh. This can happen
for example when the mesh was created with the in-house mesher \texttt{xmeshfem3D}
while using the Universal Transverse Mercator (UTM) projection but
the simulation with \texttt{xspecfem3D} was suppressing this projection
from latitude/longitude to x/y/z coordinates.

In such cases, try to change your \texttt{DATA/Par\_file} and set
e.g.:
\begin{verbatim}
SUPPRESS_UTM_PROJECTION = .false.
\end{verbatim}
to be the same in \texttt{Mesh\_Par\_file} and \texttt{Par\_file}.
This flag should be identical when using the in-house mesher \texttt{xmeshfem3D},
\texttt{xgenerate\_databases} and \texttt{xspecfem3D} together to
run simulations.

The flag determines if the coordinates you specify for your source
and station locations are given as lat/lon degrees and must be converted
to UTM coordinates. As an example, if you use \texttt{.false.} within
\texttt{Mesh\_Par\_file} then you create a mesh with \texttt{xmeshfem3D}
using the UTM projection from lat/lon as input format to UTM projected
coordinates to store the mesh point positions, which is fine. The
error then may occur if in the \texttt{Par\_file} you have this set
to \texttt{.true.} so that the \texttt{xgenerate\_databases} and \texttt{xspecfem3D}
suppress the UTM projection and assume that all coordinates you use
now for source and receiver locations are given in meters (that is,
converted) already. So it won't find the specified locations in the
used mesh. As a solutions, just change the flag in \texttt{Par\_file}
to be the same as in \texttt{Mesh\_Par\_file} and rerun \texttt{xgenerate\_databases}
and \texttt{xspecfem3D} to make sure that your simulation works fine.

@@

then change \texttt{DT} in the \texttt{DATA/Par\_file} to be somewhere
close to the maximum suggested time step. In the example above: \\
 \texttt{DT = 0.05d0} \\
 would (most probably) work fine. It could be also bigger than the
0.068 s suggested. This depends a bit on the distortions of your mesh
elements. The more regular they are, the bigger you can choose \texttt{DT}.
Just play with this value a bit and see when the simulation becomes
stable ...

>>>>

Examine the log file 'config.log'. It contains detailed informations.
In many cases, the paths to these specific compiler commands F90,
CC and MPIF90 will not be correct if `./configure` fails.\newline

@@

Make sure you're pointing to the right 'mpif90' wrapper command.\newline

Normally, this message will appear when you are mixing two different
Fortran compilers. That is, using e.g. gfortran to compile non-MPI
files and mpif90, wrapper provided for e.g. ifort, to compile MPI-files.\newline

@@

which gives you that: 27,648 / 33,792 $\sim$ 81.8 \% of all elements
are not skewed, i.e. regular elements. a fantastic value :)\newline

@@

about 6 \% of all elements have distortions with a skewness value
between 0.6 and 0.65. The skewness values give you a hint of how good
your mesh is. In an ideal world, you would want to have no distortions,
just like the 81\% from above. Those elements give you the best approximate
values by the GLL quadrature used in the spectral-element method.
However, having weakly distorted elements is still fine and the solutions
are still accurate enough. So empirically, values up to around 0.7
are tolerable, above that you should consider remeshing...\newline

@@

You will see that the mesh contains a doubling layer, where we stitch
elements together such that the size of two elements will transition
to the size of one element (very useful to keep the ratio of wavespeed
/ element\_size about constant). Those elements in this doubling layer
have higher skewness values and make up those 6 \% in the histogram.\newline

\item [{the code gives following error message "need at least one receiver":}] This
means that no stations given in the input file \texttt{DATA/STATIONS}
could be located within the dimensions of the mesh. This can happen
for example when the mesh was created with the in-house mesher \texttt{xmeshfem3D}
while using the Universal Transverse Mercator (UTM) projection but
the simulation with \texttt{xspecfem3D} was suppressing this projection
from latitude/longitude to x/y/z coordinates.\newline

In such cases, try to change your \texttt{DATA/Par\_file} and set
e.g.:
\begin{verbatim}
SUPPRESS_UTM_PROJECTION = .false.
\end{verbatim}
to be the same in \texttt{Mesh\_Par\_file} and \texttt{Par\_file}.
This flag should be identical when using the in-house mesher \texttt{xmeshfem3D},
\texttt{xgenerate\_databases} and \texttt{xspecfem3D} together to
run simulations.\newline

The flag determines if the coordinates you specify for your source
and station locations are given as lat/lon degrees and must be converted
to UTM coordinates. As an example, if you use \texttt{.false.} within
\texttt{Mesh\_Par\_file} then you create a mesh with \texttt{xmeshfem3D}
using the UTM projection from lat/lon as input format to UTM projected
coordinates to store the mesh point positions, which is fine. The
error then may occur if in the \texttt{Par\_file} you have this set
to \texttt{.true.} so that the \texttt{xgenerate\_databases} and \texttt{xspecfem3D}
suppress the UTM projection and assume that all coordinates you use
now for source and receiver locations are given in meters (that is,
converted) already. So it won't find the specified locations in the
used mesh. As a solutions, just change the flag in \texttt{Par\_file}
to be the same as in \texttt{Mesh\_Par\_file} and rerun \texttt{xgenerate\_databases}
and \texttt{xspecfem3D} to make sure that your simulation works fine.\newline

@@

then change \texttt{DT} in the \texttt{DATA/Par\_file} to be somewhere
close to the maximum suggested time step. In the example above: \newline
 \texttt{DT = 0.05d0} \newline
 would (most probably) work fine. It could be also bigger than the
0.068 s suggested. This depends a bit on the distortions of your mesh
elements. The more regular they are, the bigger you can choose \texttt{DT}.
Just play with this value a bit and see when the simulation becomes
stable ...

### 2022-11-30 doc/USER_MANUAL/notes_and_acknowledgement.tex

In order to keep the software package thread-safe in case a multithreaded
implementation of MPI is used, developers should not add modules or
common blocks to the source code but rather use regular subroutine
arguments (which can be grouped in ``derived types'' if needed for
clarity).

The Gauss-Lobatto-Legendre subroutines in \texttt{gll\_library.f90}
are based in part on software libraries from the Massachusetts Institute
of Technology, Department of Mechanical Engineering (Cambridge, Massachusetts,
USA). The non-structured global numbering software was provided by
Paul F. Fischer (Brown University, Providence, Rhode Island, USA,
now at Argonne National Laboratory, USA).

>>>>

In order to keep the software package thread-safe in case a multithreaded
implementation of MPI is used, developers should not add modules or
common blocks to the source code but rather use regular subroutine
arguments (which can be grouped in ``derived types'' if needed for
clarity).\newline

The Gauss-Lobatto-Legendre subroutines in \texttt{gll\_library.f90}
are based in part on software libraries from the Massachusetts Institute
of Technology, Department of Mechanical Engineering (Cambridge, Massachusetts,
USA). The non-structured global numbering software was provided by
Paul F. Fischer (Brown University, Providence, Rhode Island, USA,
now at Argonne National Laboratory, USA).\newline
